---
title: "Final Project: GLANCE Study"
author: "Wendy Wang and Ning Li"
date: "Tuesday, March 10, 2015"
output:
  ioslides_presentation:
    fig_caption: yes
    fig_retina: 1
    keep_md: yes
    smaller: yes
---

## Setting up some options

Let's first turn on the cache for increased performance and improved styling
```{r, cache=FALSE}
# Set some global knitr options
library("knitr")
opts_chunk$set(tidy=TRUE, tidy.opts=list(blank=FALSE, width.cutoff=60), cache=TRUE, messages=FALSE)
```

##Outline

**Reference:**

**Navarro S WE, Kantor ED, Zhang Y, Rho J, Song X, Milne GL, Lampe PD, Lampe JW. Randomized trial of glucosamine and chondroitin supplementation on inflammation and oxidative stress biomarkers and plasma proteomics profiles in healthy humans. PLoS One In press.**

- Background
- Study Design and Data Collection 
- Statistical Analysis Plan
- Data Analysis and Results
- Limitations
- Conclusions

##Background

**Glucosamine and Chondroitin (G&C)**

- One of the most popular dietary supplements in the US
- Mostly taken for osteoarthritis (OA)
- Associated with a 27-35% lower incidence of colorectal cancer, a 26-28% lower incidence of lung  cancer, 17% lower overall mortality, and a 13% lower cancer mortality.
- Safe supplements, with no known major adverse side effects.
- Regular dose
  - 1500 mg/d glucosamine hydrochloride (GHCl) + 1200 mg/d chondroitin sulfate (CS)

##Background

**Mechanisms of G&C on cancer prevention**

- Several lines of evidence from *in vitro* and preclinical studies support a possible role for G&C in reducing inflammation.
- Potentially through inhibition of nuclear factor kappa B pathway.
- No human intervention trials have been evaluated the effect of G&C on reducing inflammation or altering other pathways in healthy individuals

##Study Design

- Randomized, double-blinded, placebo-controlled crossover trial
- Ranomized on the order of treatment or placebo period
- Each intervention period lasted 28 days with a 28 day washout period in-between

##Participants

- 18 subjects: 9 males and 9 females
- Healthy, overweight(25<BMI<=32.5), non-smoking, aged 20-55 years individuals in the greater Seattle area.
- Strict exclusion criteria (e.g. medical history, medication use, large weight change, alcohol intake, supplemental use, vegetarian dietary and abnormal laboratory values)


##Data collection

- Blood samples were collected after each 28-day intervention period in the morning after a minimum of a 12-hour overnight fast.
- Blood was drawn into a tube containing EDTA for plasma. All samples were aliquoted and stored at -80°C.

##Proteomics Analysis

- Plasma samples were evaluated on a customized antibody array populated with ~3,000 full-length antibodies, printed in triplicates.
- Protein (200 µg) from a pool of albumin and IgG-depleted plasma were labeled with Cy5. Each sample was combined with reference sample labeled with Cy3.
- Unbound proteins were removed by washing and the slides scanned for Cy3 and Cy5 fluorescence in an Axon Genepix 4000B scanner.
- The Cy5/Cy3 ratio determined the relative concentration of protein compared to reference.
- Most (>85%) antibodies on the array had coefficients of variation, for triplicates, of less than 10%. Antibodies with CV>10% were excluded for further analysis.

##Array Analysis and Normalization

- The array image was scanned using a GenePix 4000B (Axon Instruments) scanner.
- For each antibody, fold-change of the signal (red channel) was compared to the reference (green channel) after background correction, the ratio noted as M value.
- Experimental variation was normalized using within-array print-tip loess and between-array quartile normalization.
- Triplicate features were summarized using their median. M values were standardized such that the mean value and standard deviation of the placebo groups were set to zero and one, respectively.
- After all processing, data were available for analysis on a total of 2938 antibodies.

##Statistical Analysis

**Overview of data**

- We have 18 treatment and 18 placebo samples
- Each has data on a 2938 antibodies
```{r}
#load packages
library(limma)
library(qvalue)

#load our data
load("Data/ma_afterNorm_organized_18pairs.Rdata")
ls()
colnames(info)

table(info$intervention)
length(rownames(ma$gene))
#Results showed 3586 antibodies. Not sure what happened?
```
- About 1100 of the antibodies are duplicates
```{r}
#Also #duplicates

```

  *Would this be of concern? (Stay tuned!)*

##Statistical Analysis Plans

**Plan 1. Replicate the reported results**

- Paired t-test using linear regression model, adjusted for batch effects and gender
- GSEA analysis using Wilcoxon Test

##Statistical Analysis Plans

**Plan 2. Alternative analysis plan**

- Averaging across duplicated antibodies
- Moderated T test using R Limma package
- GSEA using GEGG and GO pathway databases and R Camera package
- Heatmap plots of significant results

##Plan 1. Replication


##Plan 2. Alternative analysis



